High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2021-05-31 paper/main.tex
JRM edits to main text and supplement, recompiled diffs

\newcommand{\kernelshape}{S1}
\newcommand{\kernelwidth}{S2}
\newcommand{\time}{S3}
\newcommand{\features}{S4}
\newcommand{\intact}{S5}
\newcommand{\para}{S6}
\newcommand{\word}{S7}
\newcommand{\rest}{S8}
\newcommand{\pca}{S9}

@@

Overall, we found that we could reliably recover both first-order and
second-order correlations from the synthetic data
(Fig.~\ref{fig:higher_order_sims}).  When the correlations were stable
for longer intervals (Figs.~\time), or changed gradually (constant, ramping, and
event datasets), recovery performance was relatively high, and we were
better able to recover dynamic first-order correlations than
second-order correlations.  We expected that this would happen, given
that errors in our estimation procedure at lower orders necessarily
propagate to higher orders (since lower-order correlations are used to
estimate higher-order correlations).  Interestingly, we also found
that when the correlations were particularly \textit{un}stable (random
datasets), we better recovered second-order correlations.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/higher_order_sims}
  \caption{\textbf{Recovery of simulated first-order and second-order dynamic correlations.}  Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    first-order and second-order correlation matrices and the true
    (simulated) first-order and second order correlation matrices at each
    timepoint and for each synthetic dataset.  (The averages are taken across 10 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals (taken across
    datasets).  For a complete description of each synthetic dataset,
    see \textit{Synthetic data: simulating dynamic higher-order
      correlations}.  All estimates represented in this figure were
    computed using a Laplace kernel
    (width~$=20$).  \textbf{Constant correlations.}  These datasets
    have stable (unchanging) underlying second-order correlation
    matrices.  \textbf{Random correlations.} These datasets are
    generated using a new independently drawn second-order correlation
    matrix at each timepoint.  \textbf{Ramping correlations.}
    These datasets are generated by smoothly varying the underlying
    second-order correlations between the randomly drawn correlation
    matrices at the first and last timepoints.  \textbf{Event-based
      correlations.}  These datasets are each generated using five
    randomly drawn second-order correlation matrices that each remain
    stable for a fifth of the total timecourse.  The
    vertical lines denote event boundaries.}
  \label{fig:higher_order_sims}
\end{figure}

@@

The central mathematical operations in our procedure are two bookkeeping
functions, $\mathrm{vec}(\cdot)$ and $\mathrm{mat}(\cdot)$.  The
$\mathrm{vec}(\cdot)$ function takes as input a $K \times K$ symmetric
matrix and returns as output a
$\left(\frac{K^2 - K}{2} + K\right)$-dimensional column vector
containing the entries in the upper triangle and diagonal.  The
$\mathrm{mat}(\cdot)$ function inverts $\mathrm{vec}(\cdot)$ by taking
as input a $\left(\frac{K^2 - K}{2} + K\right)$-dimensional column
vector and returning a $K \times K$ symmetric matrix as output.  We
can then generate an order $n$ correlation matrix (for one timepoint, $t$) from an order
$n - 1$ template (from the same timepoint) using the Kronecker product as follows:
\begin{align}
  \mathbf{\Sigma_{n}}(t) &= \mathrm{mat}(\mathrm{vec}(\mathbf{\Sigma_{n-1}}(t)) \otimes \mathrm{vec}(\mathbf{\Sigma_{n-1}}(t))).\label{eqn:veccorr}
\end{align}

Given a timeseries of order $n$ correlation matrices, we can draw an
order $n-1$ correlation matrix for each timepoint $t$ using
\begin{align}
 \mathbf{\sigma_{n-1}}(t) &\sim \mathcal{N}\left(\mathbf{0}, \mathbf{\Sigma_n}(t)
                   \right)\label{eqn:highorder-gen1}\\
  \mathbf{\Sigma_{n-1}}(t) &= \mathrm{mat}(\mathbf{\sigma_{n-1}}(t)).\label{eqn:highorder-gen2}
  \end{align}
  We can then use repeated applications of
  Equations~\ref{eqn:highorder-gen1} and~\ref{eqn:highorder-gen2} in
  order to obtain a synthetic dataset.

 When the template first-order correlations are constructed to exhibit different
 temporal profiles (e.g., using the constant, random, ramping, and
 event procedures described above), the resulting high-order
 correlations and synthetic data will exhibit the same category of
 temporal profile. Following our approach to generating synthetic data exhibiting known
 first-order correlations, we constructed a total of 40 additional multivariate timeseries,
collectively reflecting a total of 4 qualitatively different patterns
of dynamic correlations (i.e., 10 datasets reflecting each type of
dynamic pattern: constant, random, ramping, and event).

>>>>

\newcommand{\kernelshape}{S1}
\newcommand{\kernelwidth}{S2}
\newcommand{\timeheatmap}{S3}
\newcommand{\features}{S4}
\newcommand{\intact}{S5}
\newcommand{\para}{S6}
\newcommand{\word}{S7}
\newcommand{\rest}{S8}
\newcommand{\pca}{S9}

@@

Overall, we found that we could reliably recover both first-order and
second-order correlations from the synthetic data
(Fig.~\ref{fig:higher_order_sims}).  When the correlations were stable
for longer intervals, or changed gradually (constant, ramping, and
event datasets), recovery performance was relatively high, and we were
better able to recover dynamic first-order correlations than
second-order correlations.  This is because errors in our \textit{estimation}
procedure at lower orders necessarily propagate to higher orders
(since lower-order correlations are used to estimate higher-order
correlations).  Conversely, when the correlations were particularly
\textit{un}stable (random datasets), we better recovered second-order
correlations.  This is because noise in our \textit{data generation}
procedure propagates from higher orders to lower orders (see
\textit{Synthetic data: simulating dynamic high-order correlations}).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/higher_order_sims}
  \caption{\textbf{Recovery of simulated first-order and second-order dynamic correlations.}  Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    first-order and second-order correlation matrices and the true
    (simulated) first-order and second order correlation matrices at each
    timepoint and for each synthetic dataset.  (The averages are taken across 10 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals (taken across
    datasets).  For a complete description of each synthetic dataset,
    see \textit{Synthetic data: simulating dynamic higher-order
      correlations}.  All estimates represented in this figure were
    computed using a Laplace kernel
    (width~$=20$).  \textbf{Constant.}  These datasets
    have stable (unchanging) underlying second-order correlation
    matrices.  \textbf{Random.} These datasets are
    generated using a new independently drawn second-order correlation
    matrix at each timepoint.  \textbf{Ramping.}
    These datasets are generated by smoothly varying the underlying
    second-order correlations between the randomly drawn correlation
    matrices at the first and last timepoints.  \textbf{Event.}  These datasets are each generated using five
    randomly drawn second-order correlation matrices that each remain
    stable for a fifth of the total timecourse.  The
    vertical lines denote event boundaries.}
  \label{fig:higher_order_sims}
\end{figure}

We also examined the impact of the data duration (Fig.~\timeheatmap)
and complexity (number of zero-order features; Fig.~\features) on our
ability to accurately recover ground truth first-order and
second-order dynamic correlations.  In general, we found that our approach
better recovers ground truth dynamic correlations from longer duration
timeseries data.

We also found that our approach tends to best recover data generated
using fewer zero-order features (i.e., lower complexity), although
this tendency was not strictly monotonic.  Further, because our data
generation procedure requires $\mathcal{O}(K^{4})$ memory to generate
a second-order timeseries with $K$ zero-order features, we were not
able to fully explore how the number of zero-order features affects
recovery accuracy as the number of features gets larger (e.g., as it
approaches the number of features present in the fMRI data we examine
below).  Although we were not able to formally test this to our
satisfaction, we expect that accurately estimating dynamic high-order
correlations would require data with many more zero-order features
than we were able to simulate.  Our reasoning is that
high-order correlations necessarily involve larger numbers of
lower-order features, so achieving adequate ``resolution'' high-order
timeseries might require many low-order features.

@@

The central mathematical operation in our procedure is the Kronecker
product ($\otimes$).  The Kronecker product of a $K \times K$ matrix,
$m_1$, with itself (i.e., $m_1 \otimes m_1$) produces a new
$K^2 \times K^2$ matrix, $m_2$ whose entries reflect a scaled tiling
of the entries in $m_1$.  If these tilings (scaled copies of $m_1$) are
indexed by row and column, then the tile in the
$i$\textsuperscript{th} row and $j$\textsuperscript{th} column
contains the entries of $m_1$, multiplied by $m_1(i, j)$.  Following
this pattern, the Kronecker product $m_2 \otimes m_2$ yields the
$K^4 \times K^4$ matrix $m_3$ whose tiles are scaled copies of $m_2$.
In general, repeated applications of the Kronecker self-product may be
used to generate $m_{n+1} = m_n \otimes m_n$ for $n > 1$, where
$m_{n+1}$ is a $K^{2^n} \times K^{2^n}$ matrix.  After generating a
first-order timeseries of dynamic correlations (see \textit{Synthetic
  data: simulating dynamic first-order correlations}), we use this
procedure (applied independently at each timepoint) to transform it
into a timeseries of $n$\textsuperscript{th}-order correlations.  When
$m_{n+1}$ is generated in this way, the temporal structure of the full
timeseries (i.e., constant, random, ramping, event) is preserved,
since changes in the original first-order timeseries are also
reflected in the scaled tilings of itself that comprise the
higher-order matrices.

Given a timeseries of $n$\textsuperscript{th}-order correlations, we
then need to work ``backwards'' in order to generate the order-zero
timeseries.  If the $n$\textsuperscript{th}-order correlation matrix
at a given timepoint is $m_n$, then we can generate an order $n-1$
correlation matrix (for $n > 1$) by taking a draw from
$\mathcal{N}\left(0, m_n\right)$ and reshaping the resulting vector to
have square dimensions.  Intuitively, the re-shaped matrix will look
like a noisy version of the template matrix, $m_{n-1}$.  (When
$n = 1$, no re-shaping is needed; the resulting $K$-dimensional vector
may be used as the observation at the given timepoint.)  After
independently drawing each timepoint's order $n-1$ correlation matrix
from that timepoint's order $n$ correlation matrix, this process can
be applied repeatedly until $n = 0$.  This results in a
$K$-dimensional timeseries of $T$ observations containing the
specified high-order correlations at orders 1 through $n$.  Following
our approach to generating synthetic data exhibiting known first-order
correlations, we constructed a total of 40 additional multivariate
timeseries, collectively reflecting a total of 4 qualitatively
different patterns of dynamic correlations (i.e., 10 datasets
reflecting each type of dynamic pattern: constant, random, ramping,
and event).
